Predicting Federal Emergency Management Agency Funding for U.S. Disasters

### 2021-11-12 products/manuscript/Manuscript_Disasters.Rmd
Finished modeling analysis

---
title: "Examining the Fallout Radioactivity Across Europe After the Chernobyl Meltdown"
subtitle: ""
author: Morgan Taylor
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

@@

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_ 
On April 26, 1989, a power failure at the number 4 reactor at the Chernobyl Nuclear Power Plant in the Soviet Union caused the worst nuclear disaster in history, resulting in an estimated $68 billion (adjusted for inflation) of damages. Due to the unprecedented nature of the event and the delay in notification of the general public, many questions still remain regarding the concentration and distribution of fallout radiation across several countries. As the controversy surrounding the true impact of the event remains even more than thirty years later, this study aims to better understand the distribution of the different radioisotopes from a nuclear reactor meltdown.

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
A publicly available dataset identifying sampling locations from across several European countries can be accessed [here](https://www.kaggle.com/brsdincer/chernobyl-chemical-radiation-csv-country-data).

There are nine variables included: country abbreviation, country numerical code, sampling location (city or state), latitude of sampling location, longitude of sampling location, date of sampling, and concentration of three radioisotopes (Iodine 131, Caesium 134, and Ceasium 137). There are 2051 observations included, but not all observations have complete data.

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._
There are three primary questions this study will endeavor to answer:

1. Is there a significant reduction in radioisotope concentration captured within the study period? If so, do the results represent known half-life decomposition for each radioisotope type?

2. What is the geospatial dispersion of fallout radiation captured through this publicly available sampling dataset?

3. Does the geospatial dispersion reported in the dataset match commonly accepted models for nuclear reactor meltdown? (Pending approval)

@@

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._
The raw data was obtained from [this online repository](https://www.kaggle.com/brsdincer/chernobyl-chemical-radiation-csv-country-data), but it is unclear from where the original poster obtained this data. The data were then imported into [RStudio 4.1](https://www.rstudio.com/), and the [tidyverse](https://www.tidyverse.org/) package was used to adjust variable formats and create subsets for future analyses. In lieu of using the calendar date for the sample collection, a new variable representing the number of days since the Chernobyl meltdown was created. This decision was motivated by the fact that radioactive decay is a relative measure of elapsed time rather than a calendar date. The [geosphere](https://cran.r-project.org/web/packages/geosphere/geosphere.pdf) package was used in a different feature engineering application to determine the distance between the Chernobyl nuclear power plant and each of the sampling locations. The full R code for data import and cleaning is available in the supplementary information.

>>>>

---
title: "Predicting Federal Emergency Management Agency Funding for U.S. Disasters"
subtitle: ""
author: Morgan Taylor
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

@@

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019. These laws require FEMA to publicly provide the following information for each mission assignment:

When disasters occur, one of the most complicated and difficult challenges becomes financing the response and recovery efforts. In 1988, United States (US) President Ronald Reagan signed the Robert T. Stafford Disaster Relief and Emergency Act into law to formalize the financial responsibilities of each tier of US government. Under this law, when a disaster occurs, the affected area (local or state representatives) conducts a damage assessment to estimate the amount of support required from the federal government. That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither. FEMA then deploys resources from across the federal government to support the disaster region, the volume and type of which depend upon the declaration type. However, the federal government does not necessarily incur all of the cost of these resources. Under most circumstances, this pay structure more accurately reflects a reimbursement model; that is, the states or territories submit requests to the FEMA for reimbursement after the expenses have occurred. However, there is then a reconciliation process by which the total funds a state or territory receives is often different than the original request. The accumulation of this process, including characteristics of the incident itself, are extraordinarily complex, and therefore make it difficult to predict the cost and true impact.

Moreover, the combination of rapidly increasing inflation and worsening effects of climate change has caused disasters to become more frequent and devastating on an annual basis, and the financial impacts are becoming increasingly difficult to absorb within the United States (US) government infrastructure. It would greatly benefit the federal government to be able to predict the amount of funding required for a disaster based on parameters known at the onset of the incident, as it would allow for improved funds flow and budgeting practices. Moreover, it would benefit states and territories to know which parameters increase the amount of federal government support in the aftermath of a disaster. As such, this analysis will examine the use of machine learning models in predicting FEMA expenses for a federally-declared disaster. In particular, a decision tree, bagged tree, and random forest model will be used to predict obligated and requested FEMA funds.

@@

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._
The data were obtained from the OpenFEMA Data Sets website as well as the US Department of Agriculture Economic Research Service website. The "Mission Assignment" data was cleaned to represent the requested and obligated FEMA funds for each disaster declaration for each state or territory. Feature engineering was performed to create summary variables for total number of federal agencies involved in resource assistance, the amount of funding requested by the state or territory from FEMA, the amount of funding allocated to the state or territory from FEMA, and the average cost share for FEMA of the resources provided.

The data obtained from the US Department of Agriculture Economic Research was utilized to generate state population estimates and the number of counties per state.

The data summarizing disaster declarations was cleaned to represent the unique declarations per state
